---
title: "getting some environmental covariates"
output: html_document
date: "2025-01-30"
---

extract a few environmental covariates from eDNA sampling locations on NBS 2023 survey

```{r}
library(tidyverse)
#install.packages("sf")
#packageVersion("sf")
#devtools::install_github("sean-rohan-NOAA/akgfmaps", build_vignettes = TRUE)
library(akgfmaps)
#devtools::install_github("afsc-gap-products/coldpool")
library(coldpool)
```

cold pool/ nbs_ebs temperatures 
```{r}
nbs_ebs_bt <- terra::unwrap(nbs_ebs_bottom_temperature)
nbs_ebs_st <- terra::unwrap(nbs_ebs_surface_temperature)

bt_2023 <- nbs_ebs_bt[["2023"]]
st_2023 <- nbs_ebs_st[["2023"]]

plot(bt_2023)
plot(st_2023)
```


load raster stack from AK groundfish  EFH - it is the combined environmental covariates that went into the 2023 EFH review which are single maps for each covariate based on combined data from the full time series. https://drive.google.com/drive/folders/1DHtPsnl0z0IdDb_Y30lWLBqxsoKwQ-xj. 
```{r}
library(terra)

raster.stack <- readRDS("data/raster.stack.rds")

names(raster.stack)
plot(raster.stack[[1]])
plot(raster.stack)
```

try to extract bottom temp & other covariates based on sample locations 
```{r}
sites <- read.csv("Data/NBS23_metadata.csv")%>%
  unite("site", location1, depth, sep = "_", remove = F) %>%
  select(site, longitude, latitude) %>%
  unique()

#convert site into a spatial object 
site_vect <- terra::vect(sites, geom = c("longitude", "latitude"), crs = "EPSG:4326")
plot(site_vect)

sites_reprojected <- terra::project(site_vect, crs(raster.stack))
# Check CRS
print(crs(sites_reprojected))
# Check extent
print(ext(sites_reprojected))

crs(raster.stack)
# Check CRS
print(crs(raster.stack))
# Check extent
print(ext(raster.stack))

#plot to check
btemp <- raster.stack[["btemp"]]
plot(btemp)
points(sites_reprojected, col = "red", pch = 19, cex = 1)

#extract raster values at site locations
site_values <- terra::extract(raster.stack, sites_reprojected)

#put in one df
site_enviro <- cbind(as.data.frame(sites_reprojected), site_values)
```

also extract bottom and surface temp from data in coldpool package
```{r}
plot(bt_2023)
points(sites_reprojected, col = "red", pch = 19, cex = 1)

site_bt <- terra::extract(bt_2023, sites_reprojected) %>%
  rename(bottomtemp_2023 = "2023")
site_st <- terra::extract(st_2023, sites_reprojected) %>%
  rename(surfacetemp_2023 = "2023")

site_enviro <- site_enviro %>%
  left_join(site_bt) %>%
  left_join(site_st)

site_enviro_reduced <- site_enviro %>%
  select(!coral) %>%
  select(!pen)
  
```

```{r}
write.csv(site_enviro_reduced, "outputs/NBS23_covariates.csv")
```

